---
title: "Case Study 1"
subtitle: "AKSTA Statistical Computing"
author: "Alexandra Faber, Lena Hauer, Fiona Trumshi"
date: "`r Sys.Date()`"
output: pdf_document
---

# 1. Ratio of Fibonacci numbers

## a
[solution]

## b
[solution]

## c
[solution]

# 2. Gamma function

## a

```{r}
rho <- function(n) {
  res <- gamma((n-1)/2)/(gamma(1/2) * gamma((n-2)/2))
  return(res)
}
```


## b
```{r}
rho(2000)
```

The result is NaN because R memorz limit for integer is 64-bit, meaning that higher numbers such as 2000! can not be stored and are instead defined with Inf or NaN.

## c
Using the function **lgamma** (logarithmic version of gamma) and then applying the exponential function solves the problem with larger values.

$$
\rho_n = \frac{\Gamma((n-1)/2)}{\Gamma(1/2)\Gamma((n-2)/2)} = \exp\Big(\log\frac{\Gamma((n-1)/2)}{\Gamma(1/2)\Gamma((n-2)/2)}\Big) = \exp\Big(\log(\Gamma((n-1)/2)) - \log(\Gamma(1/2))- \log(\Gamma((n-2)/2))\Big)
$$

```{r}
rhoL <- function(n){
  res <- exp(lgamma((n-1)/2) - lgamma(1/2) - lgamma((n-2)/2))
  return(res)
}
rhoL(2000)
```
```{r}
vec <- 1:2000
for (i in 1:length(vec)) {
  vec[i] <- rhoL(vec[i])
}
plot(vec)
```
The plot shows the that the implementation rhoL deals also with higher values n > 1000.

## d

```{r}
vec <- 1:2000
for (i in 1:length(vec)) {
  vec[i] <- rhoL(vec[i]) / sqrt(vec[i])
}
plot(vec)
```
The plot shows that the value exponentially increases for the first elements, while from a certain number the value slightly increases, but remains constant around the value 0.39. 
```{r}
vec[length(vec)]
```
The last value is 0.399, indicating that the limit could be 0.4, which can never be reached. 


# 3. Weak law of large numbers

## a
[solution]

## b
[solution]

## c
[solution]

## d
[solution]

## e
[solution]

## f
[solution]

# 4. Central limit theorem

## a
$$
E(\bar{X}_n) = E\Big(\frac{X_1+X_2+...+X_n}{n}\Big) = \frac{1}{n}(E(X_1)+E(X_2)+..+E(X_n)) = \frac{1}{n}n\mu = \mu
$$


## b
$$
Var(\bar{X}_n) = Var\Big(\frac{X_1+X_2+...+X_n}{n}\Big) \overbrace{=}^{X_1,...,X_n \text{ are i.i.d.}} \frac{1}{n^2}(Var(X_1)+Var(X_2)+...+Var(X_n)) = \frac{1}{n}\sigma^2
$$

## c

Simulations for $n = \{100, 1000, 10000\}$
```{r}
set.seed(2026)
sim100 <- sample(c(1:6), 100, replace = TRUE)
sim1000 <- sample(c(1:6), 1000, replace = TRUE)
sim10000 <- sample(c(1:6), 10000, replace = TRUE)
paste("Mean of 100 simulated rolls: ", mean(sim100))
paste("Mean of 100 simulated rolls: ", mean(sim1000))
paste("Mean of 100 simulated rolls: ", mean(sim10000))
```

Simulate 100 experiments and compute their means.
```{r}
set.seed(2026)
m <- 100
n <- seq(100, 10000, length.out = m)
Means <- NA

for (i in 1:m) {
  sim <- sample(c(1:6), n[i], replace = TRUE)
  Means[i] <- mean(sim)
}
plot(Means)

```
When n becomes larger the mean approaches values near to 3.5, suggesting that for bigger experiments the mean stabilizes around 3.5, while for lower n the value oscillates between 3.4 and 3.55.

```{r}
hist(Means)
```
The histogram confirms, that the mean of the means is around 3.5.  Moreover, the histogram can be approximated with a normal distribution with $\mu = 3.5$
The sample mean is more or less similar to the mean of the distribution, illustrated in the histogram.  

```{r}
paste("Sample mean: ", (1+2+3+4+5+6)/6)
```



## d

The expected value is computed as follows:

$$\mu = E[X_i] = \sum_{k=1}^{6} k \cdot P(X_i = k) = \frac{1}{6} \sum_{k=1}^{6} k = \frac{1}{6}(1+2+3+4+5+6) = \frac{21}{6} = 3.5$$

The variance is defined as:
$$
\sigma^2 = E[X_i^2] - \mu^2
$$
And the first term is computed as follows: 
$$
E[X_i^2] = \frac{1}{6} \sum_{k=1}^{6} k^2 = \frac{1}{6}(1+4+9+16+25+36) = \frac{91}{6}
$$

Then:
```{r}
sigma2 <- 91/6 - (7/2)^2
sigma2
```

$$
\sigma^2 = \frac{91}{6} - \left(\frac{7}{2}\right)^2 = \frac{91}{6} - \frac{49}{4} = \frac{182}{12} - \frac{147}{12} = \frac{35}{12} \approx 2.917
$$

## e

$$
Z_n = \frac{\bar{X}_n-\mu}{\sigma\sqrt{n}}
$$
```{r}
set.seed(2026)
mu    <- 3.5
sigma <- sqrt(35/12)
ns    <- c(100, 1000, 10000)
Ms    <- c(100, 1000, 10000)

par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))

for (M in Ms) {
  for (n in ns) {
    Zn <- replicate(M, {
      xbar <- mean(sample(1:6, n, replace = TRUE))
      (xbar - mu) / (sigma / sqrt(n))
    })
    hist(Zn,
         breaks = 30,
         freq   = FALSE,
         main   = paste0("n=", n, ", M=", M),
         xlab   = expression(Z[n]),
         col    = "steelblue",
         border = "white",
         xlim   = c(-4, 4))
    curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
  }
}
```

As n increases (from left to right) the mean of the distribution approaches the value 0, which is the mean of the standard normal distribution. When M increases (from top to bottom) the variability of the data follows the shape of a standard normal distribution ($\sigma^2 = 1$). The last combination n = 10000 M = 10000 illustrates the distribution of a standard normal distribution (in red), proving that for $n\rightarrow \infty$ the central limit theorem holds. 


# 5. Readable and efficient code

## a
[solution]

## b
[solution]

## c
[solution]

# 6. Measuring and improving performance

## a
[solution]

## b
[solution]

## c
[solution]

## d
[solution]

## e
[solution]

## f
[solution]

## g
[solution]