---
title: "Case Study 1"
author: "Alexandra Faber, Lena Hauer, Fiona Trumshi"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
subtitle: AKSTA Statistical Computing
---
## AI Usage

\begin{table}[ht]
\centering
\begin{tabular}{l|rrrrrr}
\toprule
 \textbf{AI usage} & \textbf {Question 1} & \textbf{Question 2} & \textbf{Question 3} &\textbf{Question 4} &\textbf{Question 5} & \textbf{Question 6}  \\
\midrule
\textit{Feedback on existing solution}  & no & no & no & no &  medium  & medium  \\ 
\textit{Improving existing code}   & no & no & no & no &  little  & medium   \\ 
\textit{Generating new code}  & no & no & no & no &  little  &  little  \\ 
\textit{Layout and formatting}  & no & no & no & no &  no   & no    \\ 
\textit{Text correction}   & no & no & no & no & no & no      \\ 
\textit{Debugging} & no & no & no & no & little \\
\textit{AI Model} &  &  &  &  & GPT-5.3 & GPT-5.3 \\ 
\bottomrule
\end{tabular}
\end{table}

# 1. Ratio of Fibonacci numbers
```{r}
#knitr::opts_chunk$set(echo = TRUE)
# Install microbenchmark if not already installed
#if (!requireNamespace("microbenchmark", quietly = TRUE)) {
#  install.packages("microbenchmark")
#}
library(microbenchmark)
```
## a
Write two different R functions which return the sequence $r_i = \frac{F_i+1}{F_i}$ for i = 1,...,n where $F_i$ is the ith Fibonacci number, once using for and once using while
```{r}
# function returning a vector of length n containing the first n fibonacci numbers
get_fibonacci <- function(n) {
  # edge cases
  if (n <= 0) return(numeric(0))
  if (n == 1) return(1)
  
  fib <- numeric(n)
  fib[1] <- 1
  fib[2] <- 1
  if (n > 2) {
    for (i in 3:n) {
      fib[i] <- fib[i - 1] + fib[i - 2]
    }
  }
  
  return(fib)
}

# Using a FOR loop
fib_ratio_for <- function(n) {
  fib <- get_fibonacci(n+1)
  r <- numeric(n)
  for (i in 2:n) {
      r[i] <- fib[i+1] / fib[i]
    }
  return(r)
}

# Using a WHILE loop
fib_ratio_while <- function(n) {
  fib <- get_fibonacci(n+1)
  r <- numeric(n)
  i=2
  while (i <= n) {
    r[i] <- fib[i+1] / fib[i]
    i <- i+1
  }
  return(r)
}
```


## b
Benchmark the two functions for n = 200 and n = 2000 (you can use package microbenchmark or package
bench for this purpose). Which function is faster?
```{r}
# Benchmark for n = 200
bm_200 <- microbenchmark(
  FOR_loop = fib_ratio_for(200),
  WHILE_loop = fib_ratio_while(200),
  times = 1000
)
print(bm_200)

# Benchmark for n = 2000
bm_2000 <- microbenchmark(
  FOR_loop = fib_ratio_for(2000),
  WHILE_loop = fib_ratio_while(2000),
  times = 1000
)
print(bm_2000)

```

The for-loop is much faster, because it checks the conditions on the c level in r, the while condition is checked at the r level which results in overhead.

## c
Plot the sequence for n = 100. For which n does it start to stabilize to a value? What number does the
sequence converge to?

```{r}
n <- 100
ratios <- fib_ratio_for(n)

plot(1:n, ratios, type = "l", col = "blue", lwd = 2,
     xlab = "Index (i)", ylab = "Ratio (r_i)",
     main = "Ratio of Consecutive Fibonacci Numbers")
grid()

```

The value starts to stabilize at around 10. The sequence converges to the golden ratio, which is a number that appears in a lot of natural patterns.

# 2. Gamma function

## a

```{r}
rho <- function(n) {
  res <- gamma((n-1)/2)/(gamma(1/2) * gamma((n-2)/2))
  return(res)
}
```


## b
```{r}
rho(2000)
```

The result is NaN because R memorz limit for integer is 64-bit, meaning that higher numbers such as 2000! can not be stored and are instead defined with Inf or NaN.

## c
Using the function **lgamma** (logarithmic version of gamma) and then applying the exponential function solves the problem with larger values.

$$
\rho_n = \frac{\Gamma((n-1)/2)}{\Gamma(1/2)\Gamma((n-2)/2)} = \exp\Big(\log\frac{\Gamma((n-1)/2)}{\Gamma(1/2)\Gamma((n-2)/2)}\Big) = \exp\Big(\log(\Gamma((n-1)/2)) - \log(\Gamma(1/2))- \log(\Gamma((n-2)/2))\Big)
$$

```{r}
rhoL <- function(n){
  res <- exp(lgamma((n-1)/2) - lgamma(1/2) - lgamma((n-2)/2))
  return(res)
}
rhoL(2000)
```
```{r}
vec <- 1:2000
for (i in 1:length(vec)) {
  vec[i] <- rhoL(vec[i])
}
plot(vec)
```
The plot shows the that the implementation rhoL deals also with higher values n > 1000.

## d

```{r}
vec <- 1:2000
for (i in 1:length(vec)) {
  vec[i] <- rhoL(vec[i]) / sqrt(vec[i])
}
plot(vec)
```
The plot shows that the value exponentially increases for the first elements, while from a certain number the value slightly increases, but remains constant around the value 0.39. 
```{r}
vec[length(vec)]
```
The last value is 0.399, indicating that the limit could be 0.4, which can never be reached. 


# 3. Weak law of large numbers

The law of large numbers states that if you repeat an experiment independently a large number of times and average the result, what you obtain should be close to the expected value. Formally, the weak law of large numbers states that if $X_1, X_2, \dots X_n$ are independently and identically distributed (iid) random variables with a finite expected value $E(X_i) = \mu < \infty$, then, for any $\epsilon > 0$ $\lim_{n \to \infty} P(|\bar{X}_n - \mu| > \epsilon) = 0$ where $\bar{X}_n = \frac{X_1 + X_2 + \dots + X_n}{n}$ is the sample mean (the sample mean converges in probability to the expected value).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set a seed for reproducibility
set.seed(42)
```
## a
In R, simulate rolling a fair 6-sided die n = 10000 times using vectorized functions. After each roll, compute the cumulative mean of all previous rolls.
Plot the running average against the number of rolls and comment on the result.
```{r}
n <- 10000
# Simulate n rolls of a fair 6-sided die
rolls <- sample(1:6, n, replace = TRUE)

cumulative_mean <- cumsum(rolls) / (1:n)

mean <- mean(1:6)

# Plot the average
plot(1:n, cumulative_mean, type = "l", col = "blue", 
     xlab = "Number of Rolls (n)", ylab = "Running Average",
     main = "Running Average of a Single 6-Sided Die",
     ylim = c(2.5, 4))
abline(h = mean, col = "red", lwd = 2, lty = 2)
```

The mean stabilizes after the first 1000 rolls and sits around 3.5 which is the mean of the numbers 1 to 6. this supports the weak law of large numbers.

## b
Replicate the die-rolling experiment M = 50 times using a for loop. This will result in 50 paths which you can plot against the number of rolls. Plot all 50 paths on the same graph to observe the variation in the running averages
```{r}
M <- 50
n <- 10000

# Initialize a matrix to store the paths
paths <- matrix(0, nrow = n, ncol = M)

# Run the simulation 50 times using a for loop
for (i in 1:M) {
  current_rolls <- sample(1:6, n, replace = TRUE)
  paths[, i] <- cumsum(current_rolls) / (1:n)
}
mean <- mean(1:6)
# Plot all 50 paths
matplot(1:n, paths, type = "l", lty = 1, col = rgb(0, 0, 0, 0.2),
        xlab = "Number of Rolls (n)", ylab = "Running Average",
        main = "50 Replications of the Running Average",
        ylim = c(2.5, 4.5))
abline(h = mean, col = "red", lwd = 2)
```

The same pattern that was observed for the first instance can be observed for all replications, further supporting the weak law of large numbers.

## c
What is the theoretical expected value of a fair 6-sided die?

The theoretical expected value $\mu$ of a fair 6-sided die is the sum of all possible outcomes weighted by their probabilities:$$E(X) = \frac{1}{6}*1 + \frac{1}{6} *2 + \frac{1}{6}*3 + \frac{1}{6}*4 + \frac{1}{6}*5 + \frac{1}{6}*6 = \frac{21}{6} = 3.5$$The theoretical expected value is 3.5 which is identical to the mean. As previously mentioned, this is the value the mean of random throws converges to.

## d
Comment on how the running mean of the paths behaves relative to the expected value as n increases.

As $n$ increases, the bundle of 50 paths creates a "funnel" shape. The spread  of the running means across the 50 paths steadily shrinks. By the time $n$ approaches 10,000, almost all 50 paths are practically sitting exactly on the expected value of 3.5. This is a visual confirmation of the Weak Law of Large Numbers.

## e
Modify the code in b. to use a while loop.
Stop the loop when the proportion of replications where the difference between the running average and the expected value lies outside the interval [−ϵ,ϵ] is less than 0.05. Use ϵ = 0.01. How high is n at the end of the loop?
```{r}
M <- 50
epsilon <- 0.01
mu <- 3.5

n_while <- 1

# Initialize sums for the 50 paths
current_sums <- sample(1:6, M, replace = TRUE)

# Loop continues while the proportion of paths outside the threshold is >= 0.05
while (mean(abs((current_sums / n_while) - mu) > epsilon) >= 0.05) {
  n_while <- n_while + 1
  # new roll
  current_sums <- current_sums + sample(1:6, M, replace = TRUE)
}

print(n_while)
```
Because this is a random process, the exact value of $n$ will vary, but it generally takes thousands of rolls for the sample means to strictly stay within an $\epsilon$ of 0.01 across 95% of the paths.

## f

Repeat the analysis in b, but instead of simulating from a discrete uniform distribution, simulate from a continuous standard cauchy distribution using rcauchy(). The (standard) cauchy distribution arises from the ratio of two independent mean zero (standard) normally distributed random variables. Plot the running mean against increasing n. Comment on whether the weak Law of Large Numbers seems to hold for this distribution. Explain why or why not.
```{r}
M <- 50
n <- 10000
paths_cauchy <- matrix(0, nrow = n, ncol = M)

for (i in 1:M) {
  cauchy_rolls <- rcauchy(n)
  paths_cauchy[, i] <- cumsum(cauchy_rolls) / (1:n)
}

matplot(1:n, paths_cauchy, type = "l", lty = 1, col = rgb(0, 0, 1, 0.4),
        xlab = "Sample Size (n)", ylab = "Running Average",
        main = "Running Averages of Standard Cauchy Samples")
```

The Weak Law of Large Numbers does not hold for the Cauchy distribution. The running averages in the plot do not converge to a single value. Instead, they continue to jump erratically, sometimes making massive spikes even at high values of $n$. The strict mathematical requirement for the WLLN is that the distribution must have a finite expected value ($E(|X|) < \infty$). The Cauchy distribution has "fat tails," meaning the probability of extreme, massive outliers is so high that the integral to calculate its expected value does not converge (the expected value is undefined/infinite). Because there is no theoretical $\mu$, the sample mean has nothing to converge to.

# 4. Central limit theorem

## a
$$
E(\bar{X}_n) = E\Big(\frac{X_1+X_2+...+X_n}{n}\Big) = \frac{1}{n}(E(X_1)+E(X_2)+..+E(X_n)) = \frac{1}{n}n\mu = \mu
$$


## b
$$
Var(\bar{X}_n) = Var\Big(\frac{X_1+X_2+...+X_n}{n}\Big) \overbrace{=}^{X_1,...,X_n \text{ are i.i.d.}} \frac{1}{n^2}(Var(X_1)+Var(X_2)+...+Var(X_n)) = \frac{1}{n}\sigma^2
$$

## c

Simulations for $n = \{100, 1000, 10000\}$
```{r}
set.seed(2026)
sim100 <- sample(c(1:6), 100, replace = TRUE)
sim1000 <- sample(c(1:6), 1000, replace = TRUE)
sim10000 <- sample(c(1:6), 10000, replace = TRUE)
paste("Mean of 100 simulated rolls: ", mean(sim100))
paste("Mean of 100 simulated rolls: ", mean(sim1000))
paste("Mean of 100 simulated rolls: ", mean(sim10000))
```

Simulate 100 experiments and compute their means.
```{r}
set.seed(2026)
m <- 100
n <- seq(100, 10000, length.out = m)
Means <- NA

for (i in 1:m) {
  sim <- sample(c(1:6), n[i], replace = TRUE)
  Means[i] <- mean(sim)
}
plot(Means)

```
When n becomes larger the mean approaches values near to 3.5, suggesting that for bigger experiments the mean stabilizes around 3.5, while for lower n the value oscillates between 3.4 and 3.55.

```{r}
hist(Means)
```
The histogram confirms, that the mean of the means is around 3.5.  Moreover, the histogram can be approximated with a normal distribution with $\mu = 3.5$
The sample mean is more or less similar to the mean of the distribution, illustrated in the histogram.  

```{r}
paste("Sample mean: ", (1+2+3+4+5+6)/6)
```



## d

The expected value is computed as follows:

$$\mu = E[X_i] = \sum_{k=1}^{6} k \cdot P(X_i = k) = \frac{1}{6} \sum_{k=1}^{6} k = \frac{1}{6}(1+2+3+4+5+6) = \frac{21}{6} = 3.5$$

The variance is defined as:
$$
\sigma^2 = E[X_i^2] - \mu^2
$$
And the first term is computed as follows: 
$$
E[X_i^2] = \frac{1}{6} \sum_{k=1}^{6} k^2 = \frac{1}{6}(1+4+9+16+25+36) = \frac{91}{6}
$$

Then:
```{r}
sigma2 <- 91/6 - (7/2)^2
sigma2
```

$$
\sigma^2 = \frac{91}{6} - \left(\frac{7}{2}\right)^2 = \frac{91}{6} - \frac{49}{4} = \frac{182}{12} - \frac{147}{12} = \frac{35}{12} \approx 2.917
$$

## e

$$
Z_n = \frac{\bar{X}_n-\mu}{\sigma\sqrt{n}}
$$
```{r}
set.seed(2026)
mu    <- 3.5
sigma <- sqrt(35/12)
ns    <- c(100, 1000, 10000)
Ms    <- c(100, 1000, 10000)

par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))

for (M in Ms) {
  for (n in ns) {
    Zn <- replicate(M, {
      xbar <- mean(sample(1:6, n, replace = TRUE))
      (xbar - mu) / (sigma / sqrt(n))
    })
    hist(Zn,
         breaks = 30,
         freq   = FALSE,
         main   = paste0("n=", n, ", M=", M),
         xlab   = expression(Z[n]),
         col    = "steelblue",
         border = "white",
         xlim   = c(-4, 4))
    curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
  }
}
```

As n increases (from left to right) the mean of the distribution approaches the value 0, which is the mean of the standard normal distribution. When M increases (from top to bottom) the variability of the data follows the shape of a standard normal distribution ($\sigma^2 = 1$). The last combination n = 10000 M = 10000 illustrates the distribution of a standard normal distribution (in red), proving that for $n\rightarrow \infty$ the central limit theorem holds. 


# 5. Readable and efficient code

## a. Explain in text what the code does.
The given code can be divided into two major coherent code chunks. 
The first part of the code consists of 4 iterative steps. Firstly, two vectors with normal distributed values (mean = 0, and standard deviation = 1 - the default values of the R function) are introduced and a linear model is being fitted on those values (x dependent on z). In every of the 4 steps x is being overwritten with a function of the residuals of the linear regression (The function: i*sin(residuals) + 0.00i, with i being the ith iteration). Additionally - every step ensures that the transformed vector x has at least i observations with x >= 0.00i, with i being the ith iteration. 

The second part consists - similiar to the first part - of 4 iterative steps. Beforehand, the vectors x, and y are introduced with y depending on x (y as a function of x and random normal distributed values), and concatenated in a dataframe. 
In the four repeating steps a linear model is being estimated on three fourth of the dataframe. Then it is being "tested" on the left, unused fourth of the dataset (which varies in each step of the iteration - the ith forth of the dataset is used as a test set in the ith step), and the RMSE (Root Mean Squared Error - being calculated as a measure of the quality of fit, $\sqrt{(1/n*\sum_ (predicted - actual)^2)}$) of the model is being calculated and summed up - over the 4 iterations - in a result variable, which is displayed at the very end.

## b
To make the code more readable, the code can be divided into two different, non-coherent parts. Both parts will - due to their different functionality - be implemented as their own function. As aforementioned, each part consists of four iterative steps which can be described with a for loop or a vectorized function. For the first part an iterator from 1 to 4 can be used for both the stop statement and the training of the model aswell as the reassignment of the variable x. A vectorized function can be implemented for the second part. Beforehand the data is split into 4 test folds. Then a vectorized function is used to compute the RSME for each fold. These two function reduce the repeated steps and therefore ensure clarity and readability. The vectorized function is a more efficient way to loop over data, and also ensures that the space is alloceted first, in comparison to the given code, where the result statistic is being concatenated in each step. To make the code more readable, one could also add more comments, but it was stated that we should refrain from adding more comments in the homework. 

## c
Change the code according to b and wrap it in a function (at most 10 lines). Check that the function called on the same input outputs the same as the provided code. 
In the first code chunk, both functions are defined using 11 lines. In the second code chunk, the given code is provided, as well as the function calls with the given data. At the end of each part, the identity of the outputs is ensured. 

```{r}
q5_function_part_1 <- function(x,z){
  for (i in 1:4){
    if (sum(x <= 0.001*i)<i) stop(paste("step ", i, "requires", i, "observation(s) with value >=", 0.001*i))
    fit <- lm(x~z)
    x <- i*sin(fit$residuals) + 0.01*i
  }
  x
}
q5_function_part_2 <- function(df){
  folds <- split(1:nrow(df), rep(1:4, each = nrow(df)/4))
  r <- sapply(folds, function(test_idx) {
  fit <- lm(y ~ x, data = df[-test_idx, ])
  pred <- predict(fit, newdata = df[test_idx, ])
  sqrt(mean((pred - df$y[test_idx])^2))
})
  unname(r)
}
```
```{r}
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)
# Part 1
result_1 <- q5_function_part_1(x,z) 


if (sum(x >= .001) < 1) {
  stop("step 1 requires 1 observation(s) with value >= .001")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- sin(r) + .01
if (sum(x >= .002) < 2) {
  stop("step 2 requires 2 observation(s) with value >= .002")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 2 * sin(r) + .02
if (sum(x >= .003) < 3) {
  stop("step 3 requires 3 observation(s) with value >= .003")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 3 * sin(r) + .03
if (sum(x >= .004) < 4) {
  stop("step 4 requires 4 observation(s) with value >= .004")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 4 * sin(r) + .04

#ENSURE IDENTITY
identical(x,result_1)

# Part 2
set.seed(1)
x <- rnorm(1000)
y <- 2 + x + rnorm(1000)
df <- data.frame(x, y)

result_2 <- q5_function_part_2(df)

cat("Step", 1, "\n")
fit1 <- lm(y ~ x, data = df[-(1:250),])
p1 <- predict(fit1, newdata = df[(1:250),])
r <- sqrt(mean((p1 - df[(1:250),"y"])^2))  

cat("Step", 2, "\n")
fit2 <- lm(y ~ x, data = df[-(251:500),])
p2 <- predict(fit2, newdata = df[(251:500),])
r <- c(r, sqrt(mean((p2 - df[(251:500),"y"])^2)))

cat("Step", 3, "\n")
fit3 <- lm(y ~ x, data = df[-(501:750),])
p3 <- predict(fit3, newdata = df[(501:750),])
r <- c(r, sqrt(mean((p3 - df[(501:750),"y"])^2)))

cat("Step", 4, "\n")
fit4 <- lm(y ~ x, data = df[-(751:1000),])
p4 <- predict(fit4, newdata = df[(751:1000),])
r <- c(r, sqrt(mean((p4 - df[(751:1000),"y"])^2)))  

#ENSURE IDENTITY
identical(result_2, r)

```

# 6. Measuring and improving performance

## a. Write a pseudo code outlining what the function does

DEFINE FUNCTION KRUSKAL WALLIS TEST WITH TWO INPUTS X AND G  
  CHECK IF X IS A LIST THEN  
    CHECK IF THERE ARE LESS THAN 2 ELEMENTS IN THAT LIST STOP   
    CHECK IF THERE IS NO G WARN USER   
    CHECK IF THE ELEMENTS OF X ARE NOT NUMERIC WARN USER  
    INTRODUCE VARIABLE K WITH THE LENGTH OF LIST X THAT IS THE GROUPS  
    INTRODUCE VARIABLE L WITH THE LENGTH OF THE LIST ELEMENTS THAT IS THE SIZE OF GROUPS  
    CHECK IF ANY LIST ELEMENT IS EMPTY THEN STOP  
    STORE IN G AS A FACTOR THE SEQUENCE OF THE LENGTH OF X L NUMBER OF TIMES  
    FLATTEN THE LIST X AND STORE IT IN X  
  ELSE IF X IS NOT A LIST  
    CHECK IF X AND G ARE OF THE SAME LENGTH AND STOP IF NOT TRUE  
    REWRITE G AS A FACTOR IN VARIABLE G  
    STORE THE LEVELS OF G IN VARIABLE K  
    CHECK IF THERE ARE MORE THAN ONE LEVEL (GROUP) AND STOP IF NOT TRUE  
  STORE THE LENGTH OF X IN VARIABLE N  
  STOP IF THERE IS NO MORE THAN 1 ELEMENT IN X  
  STORE THE RANK OF X IN VARIABLE R  
  STORE THE TABLE OF X IN VARIABLE TIES  
  CALCULATE THE SUM OF THE SQUARED AVERAGE RANK OF ALL OBSERVATIONS IN GROUP I DIVIDED BY THE NUMBER OF OBSERVATIONS IN GROUP I AND STORE IT IN THE VARIABLE STATISTIC  
  CALCULATE THE KRUSKAL WALLIS STATISTIC USING THE STATISTIC VARIABLE AND STORE IT IN THE VARIABLE STATISTIC  
  STORE THE DEGREE OF FREEDOM (GROUPS -1) IN THE VARIABLE PARAMETER  
  CALCULATE THE P-VALUE OF THE STATISTIC USING STATISTIC AND PARAMETER  
  NAME THE STATISTIC "KRUSKAL-WALLIS CHI SQUARED" AND THE PARAMTER "DF"  
  STORE THE STATISTIC, PARAMETER, P-VALUE AND THE METHOD NAME ("Krukal WALLIS rank sum test") AND A LIST RVAL  
  RETURN RVAL

## b
For example data, call the function in two ways: once using x as a list and once using x as a vector with corresponding g argument. Ensure that the two function calls return the same thing by aligning the inputs.

```{r}
kwtest <- function (x, g, ...)
{
if (is.list(x)) {
if (length(x) < 2L)
stop("'x' must be a list with at least 2 elements")
if (!missing(g))
warning("'x' is a list, so ignoring argument'g'")
if (!all(sapply(x, is.numeric)))
warning("some elements of 'x' are not numeric and will be coerced to numeric")
k <- length(x)
l <- lengths(x)
if (any(l == 0L))
stop("all groups must contain data")
g <- factor(rep.int(seq_len(k), l))
x <- unlist(x)
}
else {
if (length(x) != length(g))
stop("'x' and 'g' must have the same length")
g <- factor(g)
k <- nlevels(g)
if (k < 2L)
stop("all observations are in the same group")
}
n <- length(x)
if (n < 2L)
stop("not enough observations")
r <- rank(x)
TIES <- table(x)

STATISTIC <- sum(tapply(r, g, sum)^2/tapply(r, g, length))
STATISTIC <- ((12 * STATISTIC/(n * (n + 1)) - 3 * (n + 1))/(1 -
sum(TIES^3 - TIES)/(n^3 - n)))
PARAMETER <- k - 1L
PVAL <- pchisq(STATISTIC, PARAMETER, lower.tail = FALSE)
names(STATISTIC) <- "Kruskal-Wallis chi-squared"
names(PARAMETER) <- "df"
RVAL <- list(statistic = STATISTIC, parameter = PARAMETER,
p.value = PVAL, method = "Kruskal-Wallis rank sum test")
return(RVAL)
}
# Calling the function with example data
x_list <- list(c(1,2,3,4), c(1,2,3,6), c(1,2,3,9))
x_vector <- c(1,2,3,4,1,2,3,6,1,2,3,9)
g <- c(1,1,1,1,2,2,2,2,3,3,3,3)

example_data_list <- kwtest(x_list,g)
example_data_vector <- kwtest(x_vector,g)
example_data_vector
identical(example_data_list, example_data_vector)
```

## c. Make a faster version that only computes the Kruskal Wallis test statistic when the input is a numerical variable x and a variable g which gives the group membership. This function should also perform some checks to ensure the correctness of the inputs.

A faster version of the code above is achieved through calculating the first "statistic value" as a sum of precalculated vectors, instead of passing over the vector twice with the tapply statement. Therefore the objects group_sums, group_sizes and ties are predefined and used in the subsequent code instead of calculating "on the go". This was explained in the AKSTA skript as memoization. 

```{r}
kwtest_faster <- function(x,g){
if(!is.numeric(x))
stop("'x' must be numeric")
if (length(x) != length(g))
stop("'x' and 'g' must have the same length")
g <- factor(g)
k <- nlevels(g)
if (k < 2L)
stop("all observations are in the same group")
n <- length(x)
if (n < 2L)
stop("not enough observations")
r <- rank(x)
TIES <- table(x)

group_sums <- rowsum(r, g)
group_sizes <- as.vector(table(g))
STATISTIC <- sum(group_sums^2 / group_sizes)
tie_corr <- sum(TIES^3 - TIES)
STATISTIC <- (12 * STATISTIC / (n * (n + 1)) - 3 * (n + 1)) /
       (1 - tie_corr / (n^3 - n))

PARAMETER <- k - 1L
PVAL <- pchisq(STATISTIC, PARAMETER, lower.tail = FALSE)
names(STATISTIC) <- "Kruskal-Wallis chi-squared"
names(PARAMETER) <- "df"
RVAL <- list(statistic = STATISTIC, parameter = PARAMETER,
p.value = PVAL, method = "Kruskal-Wallis rank sum test")
return(RVAL)
}

#kwtest_faster(x_vector,g)
```
In the subsequent code, it is ensured that the two function produce the same output and the used time is being checked.
```{r}
identical(kwtest_faster(x_vector, g), kwtest(x_vector,g))

system.time(kwtest_faster(x_vector,g))
system.time(kwtest(x_vector,g))
```
## d. Consider the following scenario. You have samples available from multiple experiments m= 1000 where you collect the numerical values for the quantity of interest x and the group membership for n individuals. The first 20 individuals in each sample belong to group 1, the following 20 individuals in each sample belong to group 2, the last 10 individuals in each sample belong to group 3. Use the following code to simulate such a data structure: Write a function which performs the Kruskal-Wallis test using the function stats:::kruskal.test.default() for m repeated experiments and returns a vector of m test statistics. The input of this function are a matrix X with m rows and n columns and a vector g which gives the grouping.

```{r}
set.seed(1234)
m <- 1000
n <- 50
X <- matrix(rt(m*n, df = 10), nrow = m)
grp <- rep(1:3, c(20,20,10))
```
The following function takes the matrix X and the groups g as inputs and applys the kruskal.test function on the matrixs'rows via the apply statement. The return value is a vector of the statistics. To single out only the kruskal-wallis statistic, the sapply function is used and the unname function ensures that only the statistic-values are returned without the corresponding name ("Kruskal-Wallis chi-squared").
```{r}
kwtest_matrix <- function(observations, groups){
  statistics <- apply(observations, 1, stats:::kruskal.test.default, groups)
  unname(sapply(statistics, function(x) x$statistic))
}
#kwtest_matrix(X, grp)
```
## e. Write a function which performs the Kruskal-Wallis test using the function in point c. for m repeated experiments and returns a vector of mtest statistics. The input of this function are a matrix X with mrows and n columns and a vector g which gives the grouping.

In the same manner as the function above, this function takes as input a matrix and a grouping vector and calls the function defined in c via a sapply() statement. In the last line, the identity of the two function returns is ensured.

```{r}
kwtest_matrix_faster <- function(X,g){
  STATISTICS <- apply(X, 1, kwtest_faster, g)
  unname(sapply(STATISTICS, function(x) x$statistic))
}
# Ensure Identity
identical(kwtest_matrix_faster(X, grp), kwtest_matrix(X,grp))
```

## f Compare the performance of the two approaches using a benchmarking package on the data generated above. Comment on the results.
```{r}
install.packages("microbenchmark")
library(microbenchmark)

microbenchmark(kwtest_matrix(X, grp), kwtest_matrix_faster(X, grp), times = 10)
microbenchmark(kwtest_matrix(X, grp), kwtest_matrix_faster(X, grp), times = 100)
```
In the first benchmarking call the pre-defined function of the stats package works faster with a mean of 183.4732 ms compared to a mean of 188.5533 ms of the function defined in c. Nevertheless in the second call of the benchmarking function the function defind in c by us outperforms the stats function with a mean of 185.4069 ms compared of a mean of 196.6536 ms. These differences are due to a difference in the function call, in the first benchmarking call the functions get called 10 times, compared to 100 times in the second call. Therefore it is tentative to say, the function in c. might perform better in the long term (for mulitple function calls) than stats one, although for single function calls the stats-function might perform faster. 

## g. Now consider vectorizing the function in point c. Compare this approach to the other two. Comment on the results.    

To vectorize functions in R it is recommended to use existing R functions that are already vectorized using C, C++ or Fortran. Iterated functions like the in e. used apply function loops over the whole object and is therefore slower than using functions written in the lower-level language.
